Parallelizing Linear Recurrent Neural Nets Over Sequence Length

### 2017-10-27 paper/main.tex
Lots of rewriting

\begin{abstract}
Recurrent neural networks (RNNs) are widely used to model sequential data but
their non-linear dependencies between sequence elements prevent parallelizing
training over sequence length. We show the training of RNNs with only linear
sequential dependencies can be parallelized over the sequence length using the
parallel scan algorithm, leading to rapid training on long sequences with small
minibatch size.\Chris{Thought: we could rephrase this as allowing us to break
the necessity of using large batches to train fast} We develop a
parallel linear recurrence CUDA kernel and show that it can be applied to
immediately speed up several RNN architectures by up to a factor of 9. We abstract
recent work on linear RNNs into a new framework of linear surrogate RNNs and develop
a linear surrogate for the long short-term memory unit, the GILR-LSTM, which is
able to achieve throughput 30 times greater than the currently fastest-available LSTM
implementation. We show that the GILR-LSTM is able to perform well on real-world
problems,
\end{abstract}

@@

\subsection{Linear surrogate RNNs}
\label{sec:ls-rnns}
RNNs learn a transition function $s_t = f(s_{t-1}, x_t)$ which combines previous
state $s_{t-1}$ with input $x_t$ to compute current state $s_t$. Non-linear $f$
prevents application of the parallel linear recurrence algorithm and forces slow
serial evaluation. To work around this inefficiency, note that $s_t$ serves
dual purposes. In $s_t = f(s_{t-1}, x_t)$, $s_{t-1}$ serves as an input to $f$
summarizing the previous inputs while $s_t$ serves as the output of $f$ to be
passed to other layers of the network. We can decouple these uses and introduce
independent variables for each purpose: \(s_t\) is passed onto other layers of the network
and we introduce the linear surrogate \(\tilde{s}_t\) which is passed onto the
next state, with \(s_t = f(\tilde{s}_{t-1}, x_t)\). We are still able to choose a
nonlinear \(f\), our only limitation being that \(\tilde{s}_t\) must be linearly
computable.  We refer to this class of model as a linear surrogate RNN
(LS-RNN). QRNNs \cite{bradbury2017quasi} are LS-RNNs using $\tilde{h}_{t-1}
= W_k x_{t-k} + ... W_1 x_{t-1}$ and strongly typed
RNNs\cite{balduzzi2016strongly} are LS-RNNs with $\tilde{h}_t=x_{t-1}$. Although
not a rule, LS-RNNs can often be parallelized over sequence length with either
convolution or linear recurrence.

@@

\section{Experiments}
We perform several experiments, first confirming that our implementation of the
parallel linear recurrence algorithm is able to achieve higher throughput than a
similarly implemented serial version of the algorithm. We show that the parallel kernel
is up to 40 times faster than the serial equivalent for long sequence lengths and
that this speedup translates to implementations of LS-RNNs such as QRNN (where
we can get a speedup of up to 9x).

In order to illustrate that the linearization does not necessarily come at the
cost of expressibility, we show that the GILR-LSTM architecture computed with
the parallel linear recurrence algorithm, is able to outperform the optimized CuDNN LSTM
implementation
on a pathological example from the original LSTM paper
\cite{hochreiter1997long}. Finally we show that our GILR-LSTM is able to
quickly achieve state of the art results on an existing medical data
set. \Chris{I hope so\ldots}

@@

\subsubsection{Kernel performance}
We first illustrate the throughput advantage of the parallel scan algorithm for
evaluating the linear recurrence. For a minibatch comprised of \(b\) sequences
of length \(T\), we define the number of events as \(bT\) and the throughput as
the number of events per second. We implement two CUDA kernels, one which
evaluates the parallel linear recurrence described in algorithm \ref{alg:plr},
and one which evaluates the linear recurrence on GPU using the standard serial
approach (with parallelization over features but not sequence length).
This comparison is performed directly at the kernel level, avoiding
any overhead from TensorFlow. We find that the parallel kernel has
a distinct advantage at long sequence lengths with a speedup factor of up to 40x,
as shown in table \ref{table:kernel-throughput}.
At short sequence lengths, the parallel kernel does not perform well due to the
overhead of multiple passes over data and communication between processors.
\begin{table}[t]
\caption{Parallel kernel speedup on $m$ features}
\begin{center}
\begin{tabular}{@{}llll@{}}
\label{table:kernel-throughput}
Sequence Length & \(m=4\)  & \(m=32\) & \(m=128\) \\ \midrule
16              & 0.06 & 0.06 & 0.05  \\
256             & 0.22 & 0.22 & 0.86  \\
4,096           & 1.02 & 2.94 & 3.36  \\
65,536          & 38.5 & 41.8 & 17.5  \\ \bottomrule
\end{tabular}
\end{center}
\end{table}

\subsubsection{Accelerating existing RNN architectures}

Many recently introduced RNNs can be accelerated with the parallel linear recurrence
algorithm.
Table \ref{table:rnn-throughput} shows the
throughput advantage from using parallel linear recurrence compared to
serial linear recurrence. Both methods compute an identical recurrence,
so using parallel rather than serial causes no numerical changes.

We controlled for GPU memory usage by choosing minibatch size such that the product
of minibatch size and sequence length is constant. We see that the simpler architectures
(for which the linear recurrence is a higher proportion of the
computational burden) are more affected by the switch to the parallel kernel.  This is
particularly clear in the case of the QRNN, where including a wider
convolutional window results in more time spent outside of the linear recurrence and
therefore less to gain from the parallelization.
\begin{table}
\begin{center}
  \begin{tabular}{@{}lrrrr@{}}
    \label{table:rnn-throughput}
Sequence Length & GILR-LSTM & SRU & QRNN(2) & QRNN(10)\\ \midrule
16 & 0.61 & 0.28 & 0.38 & 0.78\\
256 & 0.91 & 0.84 & 0.86 & 0.99\\
4,096 & 0.98 & 1.38 & 1.18 & 1.05\\
65,536 & 1.41 & 9.21 & 6.68 & 2.05\\ \bottomrule
  \end{tabular}
\end{center}
\caption{
  Speedup obtained from using parallel linear recurrence rather than
  serial linear recurrence. We analyse this for
  SRUs \cite{lei2017} and QRNNs \cite{bradbury2017quasi} with a convolutional
  filters of size 2 and 10, and GILR-LSTM which we introduce in section \ref{sec:ls-rnns}.
  We controlled for GPU memory by setting minibatch size $b$ such that
  $bT = 65536$ for sequence length $T$.
}
\end{table}

@@

@@

\section{Conclusion}
Parallel linear recurrence can accelerate existing RNNs and be used in the design
of new architectures such as the GILR-LSTM. Parallel linear recurrence enables
rapid training with small minibatches, which further enables training models on
extremely long sequence lengths.
The impressive results in
\cite{lei2017} and other recent work comprehensively show that
linear RNNs can often be used in place of traditional non-linear RNNs.
Our results on the synthetic example continue this trend, showing that
the GILR-LSTM is able to handle sequence length regimes that are longer than any
we have seen seriously suggested in the literature. We feel that the framework
of describing linear RNNs as linear surrogates for nonlinear RNNs is a fruitful
approach, and that this, paired with the PLR algorithm, will substantially
advance the ability to train RNNs, and to understand timeseries data.

\subsubsection*{Acknowledgments}
We would like to acknowledge Kevin Bowers, Alex Meiburg, JD Co-Reyes, Carson
McNeil, Andy Palan, and several others for fruitful conversations and guidance.

>>>>

\begin{abstract}
Recurrent neural networks (RNNs) are widely used to model sequential data but
their non-linear dependencies between sequence elements prevent parallelizing
training over sequence length. We show the training of RNNs with only linear
sequential dependencies can be parallelized over the sequence length using the
parallel scan algorithm, leading to rapid training on long sequences even with small
minibatch size. We develop a parallel linear recurrence CUDA kernel and show that
it can be applied to immediately speed up training and inference of
several RNN architectures by up to 9x.
We abstract recent work on linear RNNs into a new framework of linear surrogate RNNs
and develop
a linear surrogate model for the long short-term memory unit, the GILR-LSTM, that
utilizies parallel linear recurrence.
We push the limits of sequence learning by successfully training a GILR-LSTM
on a synthetic sequence classification task with a one million timestep dependency.
SOMETHING MEDICAL?
\end{abstract}

@@

\subsection{Linear surrogate RNNs}
\label{sec:ls-rnns}
RNNs learn a transition function $s_t = f(s_{t-1}, x_t)$ which combines previous
state $s_{t-1}$ with input $x_t$ to compute current state $s_t$. Non-linear $f$
prevents application of the parallel linear recurrence algorithm and forces slow
serial evaluation. To work around this inefficiency, note that $s_t$ serves
dual purposes. In $s_t = f(s_{t-1}, x_t)$, $s_{t-1}$ serves as an input to $f$
summarizing the previous inputs while $s_t$ serves as the output of $f$ to be
passed to other layers of the network. We can decouple these uses and introduce
independent variables for each purpose: \(s_t\) is passed onto other layers of the network
and we introduce the linear surrogate \(\tilde{s}_t\) which is passed onto the
next state, with \(s_t = f(\tilde{s}_{t-1}, x_t)\). We are still able to choose a
non-linear \(f\), our only limitation being that \(\tilde{s}_t\) must be linearly
computable.  We refer to this class of model as a linear surrogate RNN
(LS-RNN). QRNNs \cite{bradbury2017quasi} are LS-RNNs using $\tilde{h}_{t-1}
= W_k x_{t-k} + ... W_1 x_{t-1}$ and strongly typed
RNNs\cite{balduzzi2016strongly} are LS-RNNs with $\tilde{h}_t=x_{t-1}$. Although
not a rule, LS-RNNs can often be parallelized over sequence length with either
convolution or linear recurrence.

@@

\section{Experiments}
We perform several experiments. First we find that our parallel linear recurrence
kernel is able to able to achieve up to 40x higher throughput than a serial implementation
when applied to long sequences. Secondly, we confirm that this kernel speedup
translates to up to a 9x speedup to LS-RNNs such as QRNNs.

In order to illustrate that the linearization does not necessarily come at the
cost of expressibility, we show that the GILR-LSTM architecture computed with
the parallel linear recurrence algorithm is able to train significantly faster
than an optimized LSTM implementation on a pathological long-term dependency
problem on a pathological example from the original LSTM paper \cite{hochreiter1997long}.
Finally we show that the GILR-LSTM is able to
quickly achieve state of the art results on an existing medical data
set. \Chris{I hope so\ldots}

@@

\subsubsection{Kernel performance}
We first illustrate the throughput advantage of the parallel scan algorithm for
evaluating the linear recurrence. For a minibatch comprised of \(b\) sequences
of length \(T\), we define the number of events as \(bT\) and the throughput as
the number of events processed per second. We implement two CUDA kernels, one which
evaluates the parallel linear recurrence described in algorithm \ref{alg:plr},
and one which evaluates the same linear recurrence on GPU in serial over sequence length
and in parallel over features and minibatch. The performance of each kernel depends on
two factors: the sequence length and the product of number of features and minibatch size.
The performance measurements for this experiment are made directly at the kernel level,
avoiding any overhead from TensorFlow. We find that the parallel kernel has
a distinct advantage at long sequence lengths with a speedup factor of up to 40x,
as shown in table \ref{table:kernel-throughput}.
The parallel kernel does not perform well at short sequence lengths due to the
overhead of multiple passes over data and communication between processors.
\begin{table}[t]
\caption{Parallel kernel speedup on $m$ features (minibatch size $= 1$)}
\begin{center}
\begin{tabular}{@{}llll@{}}
\label{table:kernel-throughput}
Sequence Length & \(m=4\)  & \(m=32\) & \(m=128\) \\ \midrule
16              & 0.06 & 0.06 & 0.05  \\
256             & 0.22 & 0.22 & 0.86  \\
4,096           & 1.02 & 2.94 & 3.36  \\
65,536          & 38.5 & 41.8 & 17.5  \\ \bottomrule
\end{tabular}
\end{center}
\end{table}

\subsubsection{Accelerating existing RNN architectures}
Several recently introduced LS-RNNs can be accelerated with the parallel linear recurrence
algorithm. We implemented SRUs, QRNNs (with filter width 2 and 10), and GILR-LSTMs
that can be computed with either the standard
serial linear recurrence algorithm or parallel linear recurrence. Both methods compute
an identical recurrence, so switching from a serial to parallel implementation
does not cause any numerical changes and takes only a single line of code changes.
Notably, both SRUs and QRNNs
claim an order of magnitude speedup compared to CuDNN LSTM when implemented with serial linear
recurrence. Any further speedup from parallel linear recurrence applies on top of the existing
speedup.
We timed train throughput (forwards and backwards propagation), but the linear time
of each pass also makes the results applicable to forwards (inference) performance. However,
parallel linear recurrence can only accelerate inference in scenarios where the entire
input sequence is known at the start of the inference phase.
We controlled for GPU memory usage within these experiments by fixing $bT = 65536$ for
minibatch size $b$ and sequence length $T$.

Table \ref{table:rnn-throughput} shows that
throughput advantage from using parallel linear recurrence compared to
serial linear recurrence reaches up to 9x. Simpler architectures
(for which the linear recurrence is a higher proportion of the
total computational load) are more affected by the switch to the parallel kernel.
This is
particularly clear in the case of the QRNN, where including wider
convolutional filters results in more time spent outside of the linear recurrence and
therefore speedup from linear recurrence parallelization.
\begin{table}[t]
\caption{Parallel kernel speedup for a variety of LS-RNNs}
\begin{center}
\begin{tabular}{@{}lrrrr@{}}
\label{table:rnn-throughput}
Sequence Length & SRU & QRNN (filter size 2) & QRNN (filter size 10) & GILR-LSTM\\ \midrule
16 & 0.28 & 0.38 & 0.78 & 0.61\\
256 & 0.84 & 0.86 & 0.99 & 0.91\\
4,096 & 1.38 & 1.18 & 1.05 & 0.98\\
65,536 & 9.21 & 6.68 & 2.05 & 1.41\\ \bottomrule
\end{tabular}
\end{center}
\end{table}

@@

Eric: I think we should delete this. We're over length, results aren't great.

@@

\section{Conclusion}
A significant portion of the success of deep learning can be attributed to massive
amounts of computation. Most of this computation is accessed through two
building blocks with highly efficient parallel implementations:
matrix multiplication and convolution. Recent research has demonstrated that
linear RNNs can achieve similar prediction accuracy to non-linear RNNs on a wide
variety of tasks in a fraction of the training time. We propose the framework of
LS-RNNs as a way to tame the growing zoo of sequential neural nets. We identify linear
recurrence as another parallelizable building block for current and future sequential models
and we use it to obtain significant speedups on already fast models.
With the power of parallel
linear recurrence we are able to solve a sequential dependency problem multiple
orders of magnitude larger than anything done prior.
Future applications of parallel linear recurrence within neural nets could include
parallel training of memory augmented models or providing a new sort of image filter
on very high resolution images.
We hope that parallel linear recurrence can be to large scale sequence
modelling what fast convolution algorithms are to image recognition.

\subsubsection*{Acknowledgments}
Acknowledgements removed for anonymous submission.
%We would like to acknowledge Kevin Bowers, Alex Meiburg, JD Co-Reyes, Carson
%McNeil, Andy Palan, and several others for fruitful conversations and guidance.
